---
output: github_document
bibliography: ./inst/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# BayesMallows

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/BayesMallows)](https://cran.r-project.org/package=BayesMallows)
[![Build Status](https://travis-ci.org/ocbe-uio/BayesMallows.svg?branch=master)](https://travis-ci.org/ocbe-uio/BayesMallows)
[![codecov](https://codecov.io/gh/ocbe-uio/BayesMallows/branch/master/graph/badge.svg)](https://codecov.io/gh/ocbe-uio/BayesMallows)

This package implements the Bayesian Mallows Model described in @vitelli2018. The user can choose between footrule, Spearman, Cayley, Hamming, Kendall, or Ulam distance. 

The following features are currently implemented:

- Complete data (@vitelli2018).

- Clustering users with similar preferences (@vitelli2018).

- Handling missing ranks by imputation (@vitelli2018).

- Handling transitive pairwise preferences by imputation (@vitelli2018).

- Estimating the partition function of the Mallows model using importance sampling (@vitelli2018) or an asymptotic approximation (@mukherjee2016).

- Non-transitive pairwise comparisons (@crispino2019).

This includes any combination thereof, e.g., clustering assessors based on pairwise preferences.

Future releases will include:

- Time-varying ranks (@asfaw2016).

- Parallelization of Markov Chains.

All feedback and suggestions are very welcome.

## Installation

To install the current release, use

```{r, eval=FALSE}
install.packages("BayesMallows")
```

To install the current development version, use

```{r, eval=FALSE}
#install.packages("devtools")
devtools::install_github("ocbe-uio/BayesMallows")
```


## References
