---
output: github_document
bibliography: ./inst/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# BayesMallows

This package implements the Bayesian Mallows Model described in @vitelli2018. At the moment, the Mallows model for complete data is implemented, and the user can choose between footrule, Spearman, Cayley, Kendall, or Hamming distance. 

Future releases will include:

- Clustering users with similar preferences (@vitelli2018).

- Handling missing ranks by imputation (@vitelli2018).

- Time-varying ranks (@asfaw2016).

- Non-transitive pairwise comparisons (@crispino2017).

All feedback and suggestions are very welcome. Feel free to create a [Pull Request](https://github.uio.no/oyss/BayesMallows/pulls) or an [Issue](https://github.uio.no/oyss/BayesMallows/issues). Please also check out the [Suggested Roadmap](https://github.uio.no/oyss/BayesMallows/wiki/Roadmap).

## Installation
To install the current development version of the package, you should clone or download this repository, and then open `BayesMallows.Rproj` in RStudio and click **Build** and then **Install and Restart** on the top menu. To get started using the package, take a look at the package documentation by running the following commands:

```{r, eval = FALSE}
library(BayesMallows)
?`BayesMallows-package`
```




## References
