---
output: github_document
bibliography: ./inst/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# BayesMallows

This package implements the Bayesian Mallows Model described in @vitelli2018. The user can choose between footrule, Spearman, Cayley, Kendall, or Hamming distance. 

The following features are currently implemented:

- Complete data (@vitelli2018).

- Clustering users with similar preferences (@vitelli2018).

- Handling missing ranks by imputation (@vitelli2018).

- Handling transitive pairwise preferences by imputation (@vitelli2018).

This includes any combination thereof, e.g., clustering assessors based on pairwise preferences.

Future releases will include:

- Time-varying ranks (@asfaw2016).

- Non-transitive pairwise comparisons (@crispino2018).

All feedback and suggestions are very welcome. Feel free to create a [Pull Request](https://github.uio.no/oyss/BayesMallows/pulls) or an [Issue](https://github.uio.no/oyss/BayesMallows/issues). Please also check out the [Suggested Roadmap](https://github.uio.no/oyss/BayesMallows/wiki/Roadmap).

## Installation
To install the current development version of the package, you should clone or download this repository, and then open `BayesMallows.Rproj` in RStudio and press CTRL + SHIFT + B. If you have trouble, please send me an [email](mailto:oystein.sorensen@medisin.uio.no). 

To get started using the package, generate its vignette with the following command:

```{r, eval=FALSE}
rmarkdown::render("./vignettes/BayesMallowsPackage.Rmd")
```



## References
