---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# BayesMallows

This is the internal repository for development of the BayesMallows package. All feedback and suggestions are very welcome. Feel free to create a [Pull Request](https://github.uio.no/oyss/BayesMallows/pulls) or an [Issue](https://github.uio.no/oyss/BayesMallows/issues).

Please also check out the [Suggested Roadmap](https://github.uio.no/oyss/BayesMallows/wiki/Roadmap).

To install the current development version of the package, you should clone or download this repository, and then open `BayesMallows.Rproj` in RStudio and click **Build** and then **Install and Restart** on the top menu. To get started using the package, take a look at the examples in the `assess_convergence` function:

```{r, eval = FALSE}
library(BayesMallows)
?assess_convergence
```


## Function Overview

This table describes the high level functions informally. It is aimed at giving an overview of the package. Check the documentation for arguments and return values.

| Function Name | Description | 
| --- | --- |
| `get_partition_function` | Returns the partition function for the Mallows model, given the necessary parameters, including distance measure. This function should have reasonable defaults for how the computation is performed. |
| `get_rank_distance` | Returns the distance between two rank vectors, in the metric specified. |
| `compute_posterior` | Runs the MCMC algorithm and computes the posterior distribution. The return values must include some convergence diagnostics. I suggest giving the return object S3 class `BayesMallows`.  |
| `assess_convergence` | Function for investigating the convergence of the MCM algorithm, and trying different parameters of the proposal distributions. |
| `plot.BayesMallows` | S3 method for plotting the posterior distribution. Note that the user only needs to call `plot`. |
| `summary.BayesMallows` | S3 method for summarizing the posterior distribution. |


## Coding Style

I follow [The tidyverse style guide](http://style.tidyverse.org/index.html). In particular, all `variable_names` with lowercase letters separated by underscore, function names are verbs, and object names are nouns.
