---
output: github_document
bibliography: ./inst/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# BayesMallows

This package implements the Bayesian Mallows Model described in @vitelli2018. At the moment, the Mallows model for complete data is implemented, and the user can choose between footrule, Spearman, Cayley, Kendall, or Hamming distance. 

Future releases will include:

- Clustering users with similar preferences (@vitelli2018).

- Handling missing ranks by imputation (@vitelli2018).

- Time-varying ranks (@asfaw2016).

- Non-transitive pairwise comparisons (@crispino2017).

All feedback and suggestions are very welcome. Feel free to create a [Pull Request](https://github.uio.no/oyss/BayesMallows/pulls) or an [Issue](https://github.uio.no/oyss/BayesMallows/issues). Please also check out the [Suggested Roadmap](https://github.uio.no/oyss/BayesMallows/wiki/Roadmap).

To install the current development version of the package, you should clone or download this repository, and then open `BayesMallows.Rproj` in RStudio and click **Build** and then **Install and Restart** on the top menu. To get started using the package, take a look at the examples in the `assess_convergence` and `compute_mallows` functions:

```{r, eval = FALSE}
library(BayesMallows)
?assess_convergence
?compute_mallows
```


## Function Overview

This table describes the high level functions informally. It is aimed at giving an overview of the package. Check the documentation for arguments and return values.

### Main Functions

| Name | Description | 
| --- | --- |
| `compute_mallows` | Runs the MCMC algorithm and computes the posterior distribution. The return values must include some convergence diagnostics. It returns an object with S3 class `BayesMallows`.  |
| `assess_convergence` | Function for investigating the convergence of the MCM algorithm, and trying different parameters of the proposal distributions. |
| `plot.BayesMallows` | S3 method for plotting the posterior distribution. Note that the user only needs to call `plot`. |
| `summary.BayesMallows` | S3 method for summarizing the posterior distribution. |

### Other Functions
The following functions are mainly used internally, but are also exported, as they may be useful by themself.

| Name | Description | 
| --- | --- |
| `get_partition_function` | Returns the partition function for the Mallows model, given the necessary parameters, including distance measure. This function should have reasonable defaults for how the computation is performed. |
| `get_rank_distance` | Returns the distance between two rank vectors, in the metric specified. |

### Data
The package comes with the potato data set (reference coming soon). This is used to illustrate the methods in the package.

| Name | Description | 
| --- | --- |
| `potato_true_ranking` | Ses `?potato_true_ranking` |
| `potato_visual` | See `?potato_visual` |
| `potato_weighing` | See `?potato_weighing` |


## References
