---
title: "Incomplete Data"
author: "Oystein Sorensen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(232312)
```

This vignette assumes that you are familiar with the topics in the introductory vignette, so make sure you read that one first. We start by describing top-k rankings, and the go on with data missing at random.

## Top-k Rankings

Imagine that the assessors in the potato experiment were asked to rank only the top-five heaviest potatoes, and that we are only interested in the top-five potatoes in our analysis. 

We generate these data by retaining only ranks 5 or higher in `potato_visual`, setting the rest to `NA`.

```{r}
library(BayesMallows)
potato_top <- ifelse(potato_visual > 5, NA_real_, potato_visual)
```


```{r, echo=FALSE, results='asis'}
knitr::kable(potato_top, caption = "Example dataset potato_top.")
```



### Excluding Leftover Items
Same MAP, but not the same posterior distributions.

## Data Missing at Random

Point: If only one item has a missing rank, this can be filled in. This is not true for top-k rankings.

# References
